Probing the internal structure of quasars using microlensing events

### 2015-08-22 paper1/discussion.tex
another set of corrections

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characteristic instances of time without making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of the parameters cannot be obtained we propose the use of a strong statistical tool such as Markov-Chain Monte Carlo. In Section 6  we have studied the possibility of identifying a crescent source and  the possibility of recovering the respective parameters. As magnification map we have used a complex numerical one generated with the microlensing code by \cite{1999A&A...346L...5W}. In addition we have added to the signal a Gaussian  noise with a SNR of 1.6.   In the experiment we have considered all nine combinations of original source profiles and assumed fitting source models. Effectively we have fitted using MCMC all three sources with all three fitting assumed models. The results of the experiment allowed us build a procedure for distinguishing the shape of the source assuming the one of the three models we have considered are a good approximation.  The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to an uniform disc and redo the fitting. Finally if the uniform disc assumption is rejected as well then the fitting should be done with a crescent source model assumption. If at each of the three steps the data rejects the model than the source cannot be approximated by none of the three models. Otherwise if at one of the steps the data does not reject the model then the respective model is a good approximation.

[Contributions from authors]\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the different source profiles discussed. Joachim Wambsganss provided the numerical code used by Manuel Rabold to create the magnification map and corresponding lighcurves used in the MCMC analysis, analysis that was performed by Irshad Mohammed in the last part of the presented work.

>>>>

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characteristic instances of time without making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of the parameters cannot be obtained we propose the use of a strong
statistical tool such as Markov-Chain Monte Carlo. In Section 6  we have studied the possibility of identifying a crescent 
source and  the possibility of recovering the respective parameters. As magnification map we have used a complex numerical 
one generated with the microlensing code by \cite{1999A&A...346L...5W}. In addition we have added to the signal a Gaussian 
noise with a SNR of 1.6.  In the experiment we have considered all nine combinations of original source profiles and assumed fitting source models.
Effectively we have fitted using MCMC all three sources with all three assumed fitting models. The results of the experiment allowed us build a 
procedure for distinguishing the shape of the source assuming that one of the three models we have considered is a good approximation.  
The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study 
the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. 
If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to 
an uniform disc and redo the fitting. Finally if the uniform disc assumption is rejected as well then the fitting should be done with a 
crescent source model assumption. If at each of the three steps the data rejects the model then the source cannot be approximated by any of the three models. 
Otherwise if at one of the steps the data does not reject the model then the respective model is a good approximation.

[Contributions from the authors]\\
P.S. provided the original idea and plan for the research project as well as multiple contributions to the analysis. M.T. simulated and studied the microlensing 
lightcurves for the different source profiles discussed for clean data and approximated fold magnification amplification. J.W. provided the training and the
numerical code used by M. R. to create the magnification map and corresponding lighcurves used in the MCMC analysis, analysis that was performed by 
I. M. in the last part of the presented work.

### 2015-08-22 paper1/theory.tex
Discussion expanded
another set of corrections

The gravitational lens equation
\begin{equation}
\vec\beta = \vec\theta - \vec\alpha(\vec\theta)
\label{eqn:lens}
\end{equation}
relates the apparent sky position $\vec\theta$ of a light source to
its true but unobservable sky position $\vec\beta$ through the bending
angle $\vec\alpha$.  The latter is an integral 
\begin{equation}
\vec\alpha = (1+z_{_L})\frac{D_S-D_L}{D_SD_L} \frac{4G}{c^2}
\int \Sigma(\vec\theta')
\frac{\vec\theta-\vec\theta'}{|\vec\theta-\vec\theta'|^2}\,
d^2\vec\theta'
\end{equation}
depending on the projected density $\Sigma(\vec\theta)$ (kg/steradian)
of the lens, the lens redshift $z_{_L}$ and the comoving distances
$D_L$ and $D_S$ to the lens and source.  The derivative of the
apparent position with respect to the source position
\begin{equation}
M(\vec\theta) =
\left(\frac{\partial\beta}{\partial\theta}\right)^{-1}
\label{eqn:magnif-matrix}
\end{equation}
is known as the magnification matrix, and its determinant
\begin{equation}
\mu(\vec\theta) = \det|M(\vec\theta)|
\end{equation}
is the brightness amplification of an image of a point source.  In
other words, the source will brighten or dim according to whether
$\mu(\vec\theta)$ is more or less than unity.  If there are multiple
mages at distinct but not observationally resolved $\vec\theta_i$ from
the same $\vec\beta$, a total brightness amplification
of\begin{equation} \mu_{\rm total} = \sum_{i} \mu(\vec\theta_i)
\end{equation}
applies. It is possible for the magnification to become formally
infinite, as a result of an eigenvalues of the magnification matrix
(\ref{eqn:magnif-matrix}) becoming infinite.  This generically happens
on curves on the $\vec\theta$ plane, known as {\em critical curves}.
Mapping a critical curve to the source plane, through the lens
equation (\ref{eqn:lens}), to the source plane $\vec\beta$ gives the
so-called caustics.  Caustics can appear in optical system, not just
gravitational lensing.  For a point source, caustics are singularities
of the magnification; for finite-size sources caustics correspond to
high and sharply changing magnification.

@@

\begin{figure}
\centering
  \includegraphics[width=.9\hsize]{plots/IRIS567_path_2.eps}
\caption{\label{fig:magnification_map} Magnification map}
\end{figure}

@@

\begin{equation}
 S_{2D}^C =  S_{2D}^{Dp} -  S_{2D}^{Dn}   (n+1)
 \label{eqn:s2d}
\end{equation}
with\\

@@

\begin{equation}
\frac{S_0^C}{2} =  \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \int_{d_c}^{R_p} \left[ \sqrt{R_p^2 - p^2} - \sqrt{R_p^2 - \left(p-c\right)^2} \Theta \left(R_p^2 - \left(p-c\right)^2 \right) \right] dp.
\end{equation}

@@

The amount of light received by an observer from a source with a gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}
which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/S1D_all.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4source_magnification.eps}
\caption{\label{fig:lightcurve_gauss} Lightcurves of crescents, Gaussian and disk for sources with identical $S_0$ and $r_{1/2}$. Crescent source has $r_{1/2}$ = 0.72754 $R_p$, $R_n$ = 0.4 $R_p$, a = 0.3/-0.3 $R_p$ }
\end{figure}

@@

\begin{equation}
 F^D(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{2 S_0^D}{ \pi R} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}} \Theta \left(R^2 - \left(p-p_s(t) \right)^2 \right) \right) \mathrm{d}p.
\end{equation}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/S1D_var_a.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4avar_magnification.eps}
\caption{\label{fig:lightcurve_disk} Lightcurves of crescent sources with $R_n$ = 0.4 $R_p$.}
\end{figure}

@@

The lightcurve of a crescent shaped source with unamplified flux $S_0^C$, radii $R_p$, $R_n$ and center displacement $a(t)$ is:
\begin{equation}
 F^c(t) = \mu_0 S_0^C + C_0 \frac{2 S_0^C}{\pi \left( R_p^2 -R_n^2 \right) } 
\left(\int_{max(0, p_s(t) - R)}^{max(0, p_s(t) + R)} \sqrt{\frac{R_p^2 - \left( p-p_s(t) \right)^2 }{p}} \mathrm{d}p 
  -  \int_{max(0, p_s(t) - a(t) - R)}^{max(0, p_s(t) -a(t) + R)} \sqrt{\frac{R_p^2 - \left( p-p_s(t) +a(t) \right)^2 }{p}}  \mathrm{d}p \right)
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/S1D_var_rn_a_poz.eps}
	\includegraphics[width = 0.48\textwidth]{plots/5Rn_back_var_magnification.eps}
\caption{\label{fig:lightcurve_crescent_back} Lightcurves of backwards crescents with  a = 0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/S1D_var_rn_a_neg.eps}
	\includegraphics[width = 0.48\textwidth]{plots/5Rn_forw_var_magnification.eps}
\caption{\label{fig:lightcurve_crescent} Lightcurve of forwards crescents with a = -0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances of a crescent source's one dimensional profile and microlensing lightcurve. }
\end{figure}

@@

\citep{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. The respective image, reproduced in figure \ref{fig:M87_image}, has been projected to a 1D profile associated to the perpendicular direction to a fold caustic approaching the image from the right. The projection is presented in the left panel of figure 10. The amplification values of the flux of light corresponding to a microlensing event are presented in the right panel of figure 10. In general the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{plots/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained from the numerical integration along the ordonata axis of the source image (figure 9) is presented in the left subplot. The corresponding lightcurve associated to the 1D profile is presented in the right subplot. The points p2 and p3 mark the start and end of the equivalent dark disk and are associated to the moments of their overlap with the caustic: t2 and t3.}
\end{figure}

>>>>

The gravitational lens equation
\begin{equation}
\vec\beta = \vec\theta - \vec\alpha(\vec\theta)
\label{eqn:lens}
\end{equation}
relates the apparent sky position $\vec\theta$ of a light source to
its true but unobservable sky position $\vec\beta$ through the bending
angle $\vec\alpha$.  The latter is an integral 
\begin{equation}
\vec\alpha = (1+z_{_L})\frac{D_S-D_L}{D_SD_L} \frac{4G}{c^2}
\int \Sigma(\vec\theta')
\frac{\vec\theta-\vec\theta'}{|\vec\theta-\vec\theta'|^2}\,
d^2\vec\theta'
\label{eqn:alpha}
\end{equation}
depending on the projected density $\Sigma(\vec\theta)$ (kg/steradian)
of the lens, the lens redshift $z_{_L}$ and the comoving distances
$D_L$ and $D_S$ to the lens and source.  The derivative of the
apparent position with respect to the source position
\begin{equation}
M(\vec\theta) =
\left(\frac{\partial\beta}{\partial\theta}\right)^{-1}
\label{eqn:magnif-matrix}
\end{equation}
is known as the magnification matrix, and its determinant
\begin{equation}
\mu(\vec\theta) = \det|M(\vec\theta)|
\end{equation}
is the brightness amplification of an image of a point source.  In
other words, the source will brighten or dim according to whether
$\mu(\vec\theta)$ is more or less than unity.  If there are multiple
mages at distinct but not observationally resolved $\vec\theta_i$ from
the same $\vec\beta$, a total brightness amplification
of\begin{equation} \mu_{\rm total} = \sum_{i} \mu(\vec\theta_i)
\end{equation}
applies. It is possible for the magnification to become formally
infinite, as a result of an eigenvalues of the magnification matrix
(\ref{eqn:magnif-matrix}) becoming infinite.  This generically happens
on curves on the $\vec\theta$ plane, known as {\em critical curves}.
Mapping a critical curve to the source plane, through the lens
equation (\ref{eqn:lens}), to the source plane $\vec\beta$ gives the
so-called caustics.  Caustics can appear in optical system, not just
gravitational lensing.  For a point source, caustics are singularities
of the magnification; for finite-size sources caustics correspond to
high and sharply changing magnification.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\hsize,bb=0 0 576 432
		]{plots/IRIS567_path_2.eps}
\caption{\label{fig:magnification_map} Magnification map used in the second part of the presented work. The white line marks the trajectory of the center of the sources for which the lightcurves presented in figure 12 are generated.}
\end{figure}

@@

\begin{equation}
 S_{2D}^C =  S_{2D}^{Dp} -  S_{2D}^{Dn}  
 \label{eqn:s2d}
\end{equation}
with\\

@@

\begin{equation}
\begin{aligned}
\frac{S_0^C}{2} & = \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \int_{d_c}^{R_p} \bigg[ \sqrt{R_p^2 - p^2} \\ 
		& - \sqrt{R_p^2 - \left(p-c\right)^2} \Theta \left(R_p^2 - \left(p-c\right)^2 \right) \bigg] dp. 
\end{aligned}
\end{equation}

@@

The amount of light received by an observer from a source with a gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}

which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth ,bb=0 0 576 520
	]{plots/S1D_all.eps}
	\includegraphics[width = 0.48\textwidth ,bb=0 0 576 520
	]{plots/4source_magnification.eps}
\caption{\label{fig:lightcurve_gauss} Lightcurves of crescents, Gaussian and disk for sources with identical $S_0$ and $r_{1/2}$. Crescent source has $r_{1/2}$ = 0.72754 $R_p$, $R_n$ = 0.4 $R_p$, a = 0.3/-0.3 $R_p$ }
\end{figure}

@@

\begin{equation}
\begin{aligned}
 F^D(t) &= \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \\
	& \bigg[ \frac{2 S_0^D}{ \pi R} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}}  \Theta \left(R^2 - \left(p-p_s(t) \right)^2 \right) \bigg] \mathrm{d}p.
\end{aligned}
\end{equation}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/S1D_var_a.eps}
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/4avar_magnification.eps}
\caption{\label{fig:lightcurve_disk} Lightcurves of crescent sources with $R_n$ = 0.4 $R_p$.}
\end{figure}

@@

The lightcurve of a crescent shaped source with unamplified flux $S_0^C$, radii $R_p$, $R_n$ and center displacement $a(t)$ is:

\begin{equation}
\begin{aligned}
 F^c(t) &= \mu_0 S_0^C + C_0 \frac{2 S_0^C}{\pi \left( R_p^2 -R_n^2 \right) } \\
	&\bigg[ \int_{max(0, p_s(t) - R)}^{max(0, p_s(t) + R)} \sqrt{\frac{R_p^2 - \left( p-p_s(t) \right)^2 }{p}} \mathrm{d}p \\
	&  -  \int_{max(0, p_s(t) - a(t) - R)}^{max(0, p_s(t) -a(t) + R)} \sqrt{\frac{R_p^2 - \left( p-p_s(t) +a(t) \right)^2 }{p}} \mathrm{d}p  \bigg] .
\end{aligned}
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/S1D_var_rn_a_poz.eps}
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/5Rn_back_var_magnification.eps}
\caption{\label{fig:lightcurve_crescent_back} Lightcurves of backwards crescents with  a = 0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/S1D_var_rn_a_neg.eps}
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/5Rn_forw_var_magnification.eps}
\caption{\label{fig:lightcurve_crescent} Lightcurve of forwards crescents with a = -0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/ch_points.eps}
        \includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances of a crescent source's one dimensional profile and microlensing lightcurve. }
\end{figure}

@@

\cite{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. The respective image, reproduced in figure \ref{fig:M87_image}, has been projected to a 1D profile associated to the perpendicular direction to a fold caustic approaching the image from the right. The projection is presented in the left panel of figure 10. The amplification values of the flux of light corresponding to a microlensing event are presented in the right panel of figure 10. In general the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth,bb=0 0 576 520
	]{plots/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained from the numerical integration along the ordonata axis of the source image (figure 9) is presented in the left subplot. The corresponding lightcurve associated to the 1D profile is presented in the right subplot. The points p2 and p3 mark the start and end of the equivalent dark disk and are associated to the moments of their overlap with the caustic: t2 and t3.}
\end{figure}
